---
title: "Extinction Threatens to Cause Morphological and Ecological Homogenization in Sharks"
title-block-banner: "#2E2D29"
title-block-banner-color: "#FFFFFF"
subtitle: "Supplementary Data:"
description: "Stanford Doer School of Sustainability, Paleobiology Lab"
date: last-modified
date-format: "YYYY-MM-DD"
mainfont: Figtree
sansfont: Georgia
footnotes-hover: true
reference-location: margin
lang: en
number-sections: false
crossref:
  chapters: true
author:
  - name: Mohamad Bazzi
    orcid: 0000-0002-9495-0781
    email: bazzi@stanford.edu
    url: https://github.com/mohabazzi
    affiliations:
      - name: Stanford University
        department: Department of Earth and Planetary Sciences
        address: CA 94305
        country: USA
        url: https://sustainability.stanford.edu/
highlight-style: github
fig-cap-location: top
lightbox: auto
format:
  stanford-quarto-html: default
other-links:
  - text: Custom Quarto Template
    icon: file-code
    href: https://github.com/mohabazzi/stanford-quarto
code-links:
  - text: Github Repo
    icon: github
    href: https://github.com/mohabazzi/carcharhinus-project
editor_options: 
  chunk_output_type: console
---

## Libraries

```{r message=FALSE, warning=FALSE, results='hide', echo=FALSE}
R.packages <- c("geomorph openxlsx dplyr ggplot2 phytools geiger phangorn tibble geometry
                 colorspace compositions Momocs ggpubr ordiBreadth tidyr akima Hmisc ggpointdensity
                 boot broom rredlist treemap ggrepel dispRity fields broom scatterplot3d")

import_packages <- 
  function(x)
  x |> trimws() |> strsplit("\\s+") |> unlist() |> 
  lapply(function(x) library(x, character.only = T))

R.packages |> import_packages()

# Custom functions.
intervals <- function(xvar,alpha = 0.975) {
  mean <- mean(xvar,na.rm = TRUE)
  std <- sd(xvar,na.rm = TRUE)
  n <- length(na.omit(xvar))
  ci.s.x <- sqrt((std^2)/n)
  pi.s.x <- std*sqrt(1 + 1/n)
  t <- qt(alpha,df = n - 1)
  ci.uci <- mean + t*ci.s.x
  ci.lci <- mean - t*ci.s.x
  pi.uci <- mean + t*pi.s.x
  pi.lci <- mean - t*pi.s.x
  res <- as.list(c(round(mean,4),round(std,4),n,round(t,4),round(ci.lci,4),round(ci.uci,4),round(pi.lci,4),round(pi.uci,4)))
  names(res) <- c('mean','st.dev','N','t.stat','Lower.CI','Upper.CI','Lower.PI','Upper.PI')
  return(res)
}
```

## Data

```{r}
# Morphometric dataset.
Data <-
  read.xlsx(xlsxFile = "Carcharhinus.xlsx") |> 
  mutate(File.Type = ".png") |> 
  # Remove C. wheeleri.
  filter(Species != "Carcharhinus wheeleri") |> 
  filter(Used == "x") # 874 specimens after removing C. wheeleri.

rownames(Data) <- paste(Data$File_Name,Data$File.Type,sep = "")

# Number of species.
carcharhinus_species <- unique(Data$Species)

# Habitat and IUCN information.
Traits <- read.xlsx(xlsxFile = "Traits.xlsx")

Traits |> 
  mutate(IUCN = factor(IUCN, levels = c("CR", "EN", "VU", "NT", "LC"))) |> 
  group_by(Size_Subclass, IUCN, Sorensen_habitat) |> 
  count() |> 
  ggplot(aes(x = IUCN, y = Size_Subclass, size = n, color = Size_Subclass)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("black", "#8F993E", "#7F2D48"),guide = "none") +
  scale_size(range = c(2, 10)) +
  labs(size = "Number of species", x = "IUCN Status", y = "Body size class") +
  facet_wrap(.~Sorensen_habitat, scales = "free") +
  theme_bw() +
  theme(aspect.ratio = 1.5,
        legend.position = "top")

# Proportions of species in each IUCN category?

# prop.table(table(Traits$IUCN)) |> 
#   as.data.frame() |> 
#   setNames(nm = c("IUCN","Proportion")) |> 
#   ggplot(mapping = aes(x = IUCN, y = Proportion)) +
#   geom_col() +
#   coord_flip() +
#   theme(aspect.ratio = 1)

# Carcharhinidae species profiles.
carcharhinidae <- read.csv2(file = "Carcharhinidae.csv",header = T)
# Plot.
carcharhinidae |>
  mutate(IUCN = factor(IUCN, levels = c("Critically Endangered","Endangered","Vulnerable","Near Threatened","Least Concern","Data Deficient"))) |> 
  group_by(Genus,IUCN) |> 
  count() |> 
  ggplot(mapping = aes(x = Genus, y = n, fill = IUCN)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values  = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50","#9E9E9E")) +
  scale_color_manual(values = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50","#9E9E9E")) +
  labs(x = "Genus", y = "Number of species", colour = "IUCN", title = "Carcharhinidae: Requiem sharks") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,hjust = 1,vjust = 1),
        axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        aspect.ratio = 1) # How many species of Carcharhinus?
```

## Landmark data

```{r}
# Custom function.
read2Dtps.noLMs <- function(file, ncurve, divide.curve = TRUE, curve.1.pts = NULL, curve.2.pts = NULL, 
                            plot = FALSE, tooth = NULL, makeSliders = TRUE, scale = FALSE) {
  require(sp)
  ignore.case = TRUE
  tpsfile <- scan(file, what = "char", sep = "\n", quiet = TRUE)
  scale.lines <- grep("SCALE=", tpsfile, ignore.case)
  lm.lines <- grep("LM=", tpsfile, ignore.case)
  n <- nspecs <- length(lm.lines)
  k <- 2
  final.results <- vector('list', 5)
  names(final.results) <- c('rawLMs', 'points.curve', 'spacedLMs', 'coords', 'sliders')
  labels <- tpsfile[grep("IMAGE=", tpsfile, ignore.case)]
  raw.LM <- vector('list', n); names(raw.LM) <- sub("IMAGE=", "", labels)
  spaced.LM <- vector('list', n); names(spaced.LM) <- sub("IMAGE=", "", labels)
  num.points <- matrix(, nrow = n, ncol = k) 

  for (i in 1:n) {
    # parse out specimens
    if (i < n) sp <- tpsfile[lm.lines[i]:(lm.lines[i+1] - 1)]
    else sp <- tpsfile[tail(lm.lines, n = 1):length(tpsfile)]
    
    # locate curve lines
    curvepts <- grep("POINTS=", sp, ignore.case = TRUE)
    print(paste("Specimen", i, "- POINTS= lines found:", length(curvepts)))
    print(sp[curvepts])  # Show the actual POINTS= lines

    if (length(curvepts) != ncurve) {
      stop(paste("Specimen", i, "does not have the expected", ncurve, "curves. Found:", length(curvepts)))
    }

    ncurvepts <- as.numeric(sub("POINTS=", "", sp[curvepts], ignore.case = TRUE))
    print(paste("Specimen", i, "- Curve point counts:", paste(ncurvepts, collapse = ", ")))
    
    num.points[i, ] <- ncurvepts
    raw.curvepts <- vector('list', 2)
    
    if (divide.curve == TRUE) {
      spaced.curvepts <- vector('list', 2)
      for (j in 1:ncurve) {
        # parse out curves
        curve <- sp[(curvepts[j] + 1):(curvepts[j] + ncurvepts[j])]
        curve <- matrix(as.numeric(unlist(strsplit(curve, "\\s+"))), ncol = k, byrow = TRUE)
        raw.curvepts[[j]] <- curve
        if (j == 1) {
          L.lands <- Line(curve)
          semi.lands <- spsample(L.lands, curve.1.pts, type = "regular", offset = c(0,1))@coords
          spaced.curvepts[[j]] <- semi.lands
        }
        if (j == 2) {
          curve <- curve[ncurvepts[j]:1, ]
          L.lands <- Line(curve)
          semi.lands <- spsample(L.lands, curve.2.pts - 1, type = "regular", offset = c(0,1))@coords
          spaced.curvepts[[j]] <- rbind(semi.lands, curve[ncurvepts[j], ])
        }
      }
      spaced.LM[[i]] <- rbind(spaced.curvepts[[1]], spaced.curvepts[[2]])
      raw.LM[[i]] <- rbind(raw.curvepts[[1]], raw.curvepts[[2]][-1, ])
      p <- nrow(rbind(spaced.curvepts[[1]], spaced.curvepts[[2]]))
    }

    if (divide.curve == FALSE) {
      for (j in 1:ncurve) {
        curve <- sp[(curvepts[j] + 1):(curvepts[j] + ncurvepts[j])]
        curve <- matrix(as.numeric(unlist(strsplit(curve, "\\s+"))), ncol = k, byrow = TRUE)
        raw.curvepts[[j]] <- curve
      }
      L.lands <- Line(rbind(raw.curvepts[[1]], raw.curvepts[[2]][-1, ]))
      semi.lands <- spsample(L.lands, curve.1.pts, type = "regular")@coords
      spaced.LM[[i]] <- semi.lands
      raw.LM[[i]] <- rbind(raw.curvepts[[1]], raw.curvepts[[2]][-1, ])
      p <- curve.1.pts
    }
  }

  coords <- array(unlist(spaced.LM), c(p, k, n), dimnames = list(NULL, NULL, names(spaced.LM)))

  if (scale) {
    imscale <- as.numeric(sub("SCALE=", "", tpsfile[grep("SCALE", tpsfile, ignore.case)], ignore.case))
    imscale <- aperm(array(rep(imscale, p * k), c(n, k, p)), c(3, 2, 1))
    coords <- coords * imscale
  }

  final.results$rawLMs <- raw.LM
  final.results$points.curve <- num.points
  final.results$spacedLMs <- spaced.LM
  final.results$coords <- coords

  if (plot == TRUE) {
    par(mfrow = c(1, 2))
    plot(raw.LM[[tooth]], xlab = NA, ylab = NA, main = "Digitized Points")
    plot(coords[,,tooth], xlab = NA, ylab = NA, main = "Resampled & Scaled Points")
  }

  if (makeSliders == TRUE) {
    final.results$sliders <- data.frame(before = 1:(p-2), slider = 2:(p-1), after = 3:p)
  }

  return(final.results)
}
```

<!--Import shape data-->

```{r results='hide',echo=TRUE,message=FALSE,warning=FALSE}
# Landmark data.
LMs <- read2Dtps.noLMs(file = "Images/landmarks.TPS", ncurve = 2,
                       divide.curve = TRUE,
                       curve.1.pts = 80, curve.2.pts = 80,
                       scale = TRUE)
```

## Phylogeny

```{r}
# MCC tree of sharks.
tree <- readNexus(file = "mcc_tree.nex")

species_tips <- 
  as.data.frame(grep("Carcharhinus",tree$tip.label,value = T)) |> 
  setNames(nm = "species") |> 
  filter(species %in% gsub(" ","_",carcharhinus_species)) |> 
  column_to_rownames(var = "species") # 30 species.
  
carcharhinus_tree <- treedata(phy = tree,data = species_tips, warnings = FALSE)$phy

# Carcharhinidae tree.
carcharhinidae_tree <- treedata(phy = tree,
                                data = carcharhinidae |> mutate(Species = gsub(" ","_",Species)) |> column_to_rownames(var = "Species"),
                                warnings = FALSE)$phy

plotTree(ladderize(carcharhinidae_tree),ftype="i",lwd=1,tip.label=TRUE,fsize=.5, edge.color = "black")
```

## Diets and niche breadth

Diets for *Carcharhinus leiodon*, *Carcharhinus hemiodon* and *Carcharhinus borneensis* were not included in the dataset by Corets 1999.

```{r}
# Diets.
Diets <- read.xlsx(xlsxFile = "Diets.xlsx")
rownames(Diets) <- paste(Diets$Genus,Diets$Species,sep = "_")

Diets$IUCN_2024 <- factor(Diets$IUCN_2024, levels = c("CR","EN","VU","NT","LC"))

# Subset tree.
tree_diets <- geiger::treedata(phy = carcharhinus_tree,data = Diets, warnings = FALSE)$phy # 27 species with diet and tree availability.

# Plot.
graphics::layout(matrix(1:1,nrow = 1),respect = T)

phylo.heatmap(tree = tree_diets, X = Diets[,6:15], standardize = F,
              colors = rev(sequential_hcl(11,palette = "Red-Yellow")),split=c(0.6,0.4),
              fsize = .5) # Overwheling fish-eaters across the board.

# Compositional diets.
closed_diets <- clo(X = Diets[,6:15])

# Convert to binary matrix.
binary_niche <- 
  Diets |>
  select(6:15) |> 
  mutate(across(everything(), ~ ./100)) |> 
  mutate_all(~ ifelse(. == 0.0000000000, 0, 1))
```

<!-- Niche breadth analysis-->

```{r warning=FALSE, message=FALSE}
# ODB analysis.
odb <- ordi.breadth(dat = binary_niche,dist.method = "jaccard")

# Data frame.
odb_results <- 
  data.frame(Axis = odb$tot.breadth, Index = 1:27,
             Species = odb$species,
             Genera = gsub("_.+$","", odb$species)
             )

odb_results <-
  odb_results |> 
  left_join(Traits |> mutate(Species = as.factor(paste(Genus,Species, sep = "_"))) |> 
              select(Species,IUCN,Habitat,Secondary_Habitat,Size_Subclass,Ecomorphotype,Sorensen_habitat), by = "Species"
            ) # 27 species.

odb_results <-
  odb_results |> 
  as.data.frame() |> 
  mutate(IUCN = factor(IUCN, levels = c("CR","EN","VU","NT","LC"))) |> 
  mutate(Size_Subclass = factor(Size_Subclass, levels = c("small_shark","medium_shark","large_shark")))

# IUCN colours.
iucn_categories <- factor(setNames(Diets$IUCN_2024, paste(Diets$Genus,Diets$Species, sep = "_")))
iucn_cols <- setNames(c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50"), 
                      levels(iucn_categories))

# Plot.
p1 <-
  odb_results |> 
  mutate(abbreviated_species = Species) |> 
  mutate(abbreviated_species = gsub("Carcharhinus_","C. ",abbreviated_species)) |> 
  ggplot(mapping = aes(x = Index, y = Axis, size = Size_Subclass,colour = IUCN, shape = Sorensen_habitat)) +
  geom_point() +
  scale_fill_manual(values  = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  scale_color_manual(values = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  scale_x_continuous(limits = c(0,30)) +
  labs(x = "Species", y = "Total diet breadth", colour = "IUCN", size = "Body size class",shape = "Habitat") +
  # Quantiles.
  geom_hline(yintercept = c(quantile(odb_results$Axis, c(0.33, 0.66))), 
           linetype = "dashed",linewidth = 1, color = "grey50") +
  annotate("text", x = 0.5, y = quantile(odb_results$Axis, 0.15), 
         label = "Specialist", hjust = 0) +
  annotate("text", x = 0.5, y = quantile(odb_results$Axis, 0.85), 
         label = "Generalist", hjust = 0) +
  theme_bw() +
  theme(axis.title = element_text(face = "bold"),
         legend.position = "right",
         aspect.ratio = 1)

# p1 |> print()

# Raw diets.
p2 <- 
  Diets |> 
  select(6:15) |> 
  as.data.frame() |> 
  rownames_to_column(var = "taxa") |> 
  reshape::melt() |> 
  ggplot(aes(x= "", y=value, fill=variable)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_fill_manual(values = colorspace::qualitative_hcl(n = 10,palette = "Dark 3")) +
  labs(fill = NULL) +
  coord_polar("y", start=0) +
  theme_void() +
  facet_wrap(.~taxa) +
  theme(plot.title = element_text(size = 3),
        strip.text = element_text(size = 7))

# p2 |> print()
```

## Procrustes superimposition

```{r message=FALSE}
# Slider file.
slider_file <-
  rbind(define.sliders(c(1,2:79,80)),
        define.sliders(c(80,81:159,160))
        )

# Procrustes superimposition.
tooth_gpa <- gpagen(A = LMs$coord[,,rownames(Data)],curves = slider_file,
                    approxBE = T,
                    Proj = T,
                    print.progress = F,PrinAxes = T) # Without C. wheeleri.

centroid_size <- tooth_gpa$Csize

# tooth_gpa$coords |> dim() # 874 specimens.
```

```{r}
# Check that the correct was made.
p3 <-
  dimnames(tooth_gpa$coords)[[3]] |> 
  as.data.frame() |> 
  setNames(nm = "Species") |> 
  mutate(Species = gsub("-"," ",Species)) |> 
  mutate(Species = gsub("_.*","",Species)) |> 
  group_by(Species) |> 
  count() |> 
  ggplot(mapping = aes(x = Species, y = n)) +
  geom_col() +
  coord_flip() +
  theme(aspect.ratio = 1)

# print(p3)

# Size.
names(tooth_gpa$Csize) <- gsub(pattern = "_.*",replacement = "",x = names(tooth_gpa$Csize))

# Upper versus Lower teeth.
p4 <- 
  Data |> 
  group_by(Species,Jaw) |> 
  count() |> 
  ggplot(mapping = aes(x = Species,y = n, fill = Jaw)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0,60)) +
  # scale_colour_manual(values = c("#AD9490","lightblue")) +
  scale_fill_manual(values =   c("#AD9490","lightblue")) +
  coord_flip() +
  labs(x = NULL, y = "Sample size") +
  theme_bw() +
   theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, face = "italic"),
        axis.title = element_text(face = "bold"),
        # legend.position = c(0.8,0.8),
        aspect.ratio = 2)

# p4 |> print()

species_sample <-
  data.frame(species = names(tooth_gpa$Csize),
               cz = as.vector(log10(tooth_gpa$Csize))
               ) |> 
  group_by(species) |> 
  count() |> 
  mutate(LA = paste(species,"n=",n,sep = " ")) |> 
  mutate(LA = gsub("= ","=",LA)) |> 
  mutate(LA = gsub("n=","\nn=",LA)) |> 
  mutate(LA = gsub("-"," ",LA))

label_vec <- setNames(species_sample$LA, gsub("-"," ",species_sample$species))

p5 <-
  data.frame(species = gsub("-"," ",names(tooth_gpa$Csize)),
             cz = as.vector(log10(tooth_gpa$Csize)),
             jaw = gsub(".*(_[LU]).*", "\\1", names(centroid_size))
             ) |>
  mutate(jaw = gsub("_L", "Lower", jaw)) |> 
  mutate(jaw = gsub("_U", "Upper", jaw)) |> 
  ggplot(mapping = aes(x = cz, y = species, fill = jaw)) +
  geom_boxplot() +
  scale_fill_manual(values =  c("#AD9490","lightblue")) +
  scale_color_manual(values = c("#AD9490","lightblue")) +
  labs(x = "Log-transformed centroid size",y = NULL, fill = "Jaw unit") +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        axis.title = element_text(face = "bold"),
        aspect.ratio = 2) +
  scale_y_discrete(labels = label_vec)

# p5 |> print()
```

## Principal components analysis

### Specimen level

```{r}
# PCA
pca <- gm.prcomp(A = tooth_gpa$coords)

# Variance explained.
# round(unlist(pca$sdev^2/sum(pca$sdev^2))[1:4],digits = 4) * 100 # PC1 = 72.0%, PC2 = 14.81%, PC3 = 7.17%

pca_data <-
  data.frame(PC1 = pca$x[,1],
             PC2 = pca$x[,2],
             PC3 = pca$x[,3],
             Species = gsub("-"," ", x = gsub("_.*","",x = rownames(pca$x))),
             Jaw = gsub("_L","Lower",gsub("_U","Upper", gsub(".*(_L|_U).*", "\\1", rownames(pca$x)))),
             Serrations = setNames(Data$Serration,rownames(Data)),
             cz = as.numeric(unname(tooth_gpa$Csize))
             ) |> 
  left_join(Traits |> 
              mutate(Species = paste(Genus,Species,sep = " ")) |>
              select(Species, IUCN, Habitat,Secondary_Habitat,Size_Subclass,Ecomorphotype,Sorensen_habitat),by = "Species") |> 
  mutate(IUCN = factor(IUCN, levels = c("CR","EN","VU","NT","LC"))) |> 
  # Adjust serration factor.
  mutate(Serrations = stringr::str_squish(Serrations))

rownames(pca_data) <- rownames(pca$x)

# Species-averages.
average_species <-
  pca_data |> 
  group_by(Species) |> 
  summarise(PC1 = mean(PC1),PC2 = mean(PC2),PC3 = mean(PC3)) |> 
  pivot_longer(cols = c(2:4),names_to = "variable")

# Species
specimen_axes <-
  pca_data |> 
  pivot_longer(cols = c(1:3),names_to = "variable")

# Plot specimen level PCA.
specimen_axes |>
  ggplot(mapping = aes(x = Species, y = value, colour = Species)) +
  geom_boxplot(colour = "#AD9490", fill = "#E2DED4",show.legend = F) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar",width = .10, lwd = 1, linetype = 1, col = "#2E2D29") +
  geom_point(mapping = aes(x = Species, y = value),data = average_species,size = 2,show.legend = F,
             fill = "#FFFFFF", colour = "#2E2D29", shape = 21) +
  geom_hline(yintercept = 0.0,linewidth = 1,linetype = 2,colour = "black") +
  coord_flip() +
  # Add IUCN status.
  geom_point(data = specimen_axes |> filter(variable == "PC1"), 
             aes(y = -0.5, x = Species, fill = IUCN), 
             inherit.aes = FALSE,
             shape = 21, colour = "#2E2D29", size = 5,
             show.legend = F) +
  # geom_text(data = specimen_axes |> filter(variable == "PC1"), 
  #           aes(y = -0.5, x = Species, label = IUCN), 
  #           inherit.aes = FALSE,
  #           hjust = 0.5, vjust = 0.5, size = 2) +
  # Custom colors.
  scale_fill_manual(values  = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  facet_wrap(.~variable, scales = "free_x",
             labeller = labeller(variable = c(
               PC1 = "Principal Component 1 (72.0%%)",
               PC2 = "Principal Component 2 (14.81%)",
               PC3 = "Principal Component 3 (7.17%)"))) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        aspect.ratio = 2,
        strip.text = element_text(face = "bold"))

average_iucn <-
  pca_data |> 
  group_by(IUCN) |> 
  summarise(PC1 = mean(PC1),PC2 = mean(PC2),PC3 = mean(PC3))

# Plot by jaw: PC1
p6 <- pca_data |> 
  ggplot(mapping = aes(x = Species, y = PC1, colour = Jaw)) +
  geom_boxplot() +
  scale_fill_manual(values =  c("#AD9490","lightblue")) +
  scale_color_manual(values = c("#AD9490","lightblue")) +
  geom_hline(yintercept = 0.0,linetype = 2) +
  labs(y = "Principal Component 1 (72.0%)", x = NULL) +
  coord_flip() +
  theme_bw() +
  theme(aspect.ratio = 2,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right",
        axis.text.y = element_text(face = "italic"))

# PC2.
p7 <- pca_data |> 
  ggplot(mapping = aes(x = Species, y = PC2, colour = Jaw)) +
  geom_boxplot() +
  scale_fill_manual(values =  c("#AD9490","lightblue")) +
  scale_color_manual(values = c("#AD9490","lightblue")) +
  geom_hline(yintercept = 0.0,linetype = 2) +
  labs(y = "Principal Component 2 (14.81%)", x = NULL) +
  coord_flip() +
  theme_bw() +
  theme(aspect.ratio = 2,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right",
        axis.text.y = element_text(face = "italic"))

# IUCN
p8 <- pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = IUCN)) +
  geom_point() +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = IUCN),data = average_iucn,size = 6) +
  scale_fill_manual(values  = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  scale_color_manual(values = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right")

highlight_data <- 
  pca_data |> 
  group_by(IUCN) |> 
  mutate(highlight_group = IUCN) |> 
  ungroup()

# Plot
p9 <- ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2),
             colour = "#D0D0D0", size = 1) +
  geom_point(data = highlight_data, 
             aes(x = PC1, y = PC2), size = 2) +
  facet_wrap(~ highlight_group) +
  scale_colour_manual(values = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  scale_fill_manual(values =   c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none", aspect.ratio = 1)

# Jaws and serrations.
p10 <- 
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Serrations)) +
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = qualitative_hcl(n = 4,"Dark 3")) +
  scale_colour_manual(values = qualitative_hcl(n = 4,"Dark 3")) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)", colour = "Serration Type") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "none")

# Ecomorphotype.
p11 <- 
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Ecomorphotype)) +
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values =   c("#544948","#20d420")) +
  scale_colour_manual(values = c("#544948","#20d420")) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "none")

average_body_class <- 
  pca_data |> 
  group_by(Size_Subclass) |> 
  summarise(PC1 = mean(PC1),PC2 = mean(PC2),PC3 = mean(PC3))

p12 <- 
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Size_Subclass)) +
  geom_point() +
  # geom_point(mapping = aes(x = PC1,y = PC2, colour = Size_Subclass),data = average_body_class, size = 6) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values =   c("black","#8F993E","#7F2D48")) +
  scale_colour_manual(values = c("black","#8F993E","#7F2D48")) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)", colour = "Body size class") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "none")

# Main habitat.
p13 <-
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Habitat)) +
  geom_point() +
  scale_fill_manual(values  = c("#007C92","#B83A4B")) +
  scale_color_manual(values = c("#007C92","#B83A4B")) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right")

# Secondary habitat.
p14 <- 
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Secondary_Habitat)) +
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  scale_colour_manual(values = c("#FFC20A", "#0C7BDC")) +
  scale_fill_manual(values =   c("#FFC20A", "#0C7BDC")) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right")

# Tertiary habitat.
p15 <- 
  pca_data |> 
  ggplot(mapping = aes(x = PC1, y = PC2, colour = Sorensen_habitat)) +
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values =   c("#00AFBB","#E7B800", "#FC4E07")) +
  scale_colour_manual(values = c("#00AFBB","#E7B800", "#FC4E07")) +
  labs(x = "Principal Component 1 (72.0%)", y = "Principal Component 2 (14.81%)", colour = "Habitat") +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right")
```

<!--TPS grids -->

```{r results='hide',echo=FALSE,fig.show='hide'}
# Morphologies along PC-axes
plotRefToTarget(pca$shapes$shapes.comp1$min,pca$shapes$shapes.comp1$min)
plotRefToTarget(pca$shapes$shapes.comp1$max,pca$shapes$shapes.comp1$max)
plotRefToTarget(pca$shapes$shapes.comp2$min,pca$shapes$shapes.comp2$min)
plotRefToTarget(pca$shapes$shapes.comp2$max,pca$shapes$shapes.comp2$max)
plotRefToTarget(pca$shapes$shapes.comp3$min,pca$shapes$shapes.comp3$min)
plotRefToTarget(pca$shapes$shapes.comp3$max,pca$shapes$shapes.comp3$max)
```

### Species level analysis

<!-- Average morphology -->

Are smaller, medium or larger species at more risk of extinction? Are threatned species living in extreme areas of morphospace? 

```{r}
species_coords <- tooth_gpa$coords

dimnames(species_coords)[[3]] <- gsub("-"," ",dimnames(species_coords)[[3]])
dimnames(species_coords)[[3]] <- gsub("_.*", "",dimnames(species_coords)[[3]])
dimnames(species_coords)[[3]] <- gsub(" ", "_",dimnames(species_coords)[[3]])

# Computing an average shape.
species_subset <- coords.subset(A = species_coords, group = as.factor(dimnames(species_coords)[[3]]))

species_subset <- 
  lapply(species_subset, mshape) |>
  simplify2array()

# Shape analysis via PCA.
average_pca <- gm.prcomp(A = species_subset,phy = carcharhinus_tree)
# Variance explained.
barplot(round(unlist(average_pca$sdev^2/sum(average_pca$sdev^2)),4)* 100,las = 2,horiz = F,ylab = "Variance Explained")

# Plot tps grids.
layout(matrix(1:4,nrow = 2,byrow = T),respect = T)

plotRefToTarget(average_pca$shapes$shapes.comp1$min,average_pca$shapes$shapes.comp1$min,mag = 3)
plotRefToTarget(average_pca$shapes$shapes.comp1$max,average_pca$shapes$shapes.comp1$max,mag = 3)
plotRefToTarget(average_pca$shapes$shapes.comp2$min,average_pca$shapes$shapes.comp2$min,mag = 3)
plotRefToTarget(average_pca$shapes$shapes.comp2$max,average_pca$shapes$shapes.comp2$max,mag = 3)

par(mfrow = c(1,2))
plotRefToTarget(average_pca$shapes$shapes.comp3$min,average_pca$shapes$shapes.comp3$min,mag = 3) # Less telling 
plotRefToTarget(average_pca$shapes$shapes.comp3$max,average_pca$shapes$shapes.comp3$max,mag = 3) # Less telling 

# IUCN colours for all species and not only those with diets.
iucn_species <- factor(setNames(Traits$IUCN, paste(Traits$Genus,Traits$Species, sep = "_")),
                       levels = c("CR","EN","VU","NT","LC")) # 30 taxa
iucn_colours <- setNames(c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50"), 
                      levels(iucn_species))
# Ecomorphology.
ecomorph_species <- factor(setNames(Traits$Ecomorphotype, paste(Traits$Genus,Traits$Species, sep = "_")))
ecomorph_colours <- setNames(c("#544948","#417865"),levels(ecomorph_species))

# Main Habitats.
habitat_species <- factor(setNames(Traits$Habitat, paste(Traits$Genus,Traits$Species, sep = "_")))
habitat_colours <- setNames(c("#766253","#016895"),levels(habitat_species))

# Secondary habitat.
secondary_habitat_species <- factor(setNames(Traits$Secondary_Habitat, paste(Traits$Genus,Traits$Species, sep = "_")))
secondary_colours <- setNames(c("#EDE8D0","black"),levels(secondary_habitat_species))

# Tertiary habitat.
sorenson_habitats <- factor(setNames(Traits$Sorensen_habitat, paste(Traits$Genus,Traits$Species, sep = "_")))

# Body size class
size_species <- factor(setNames(Traits$Size_Subclass, paste(Traits$Genus,Traits$Species, sep = "_")))
size_colours <- setNames(c("#2D716F","#766253","#A6B168"),levels(size_species))

# Serrations.
serration_type <- 
  Data |> 
  select(Species,Serration) |> 
  distinct(Species,.keep_all = T) |> 
  mutate(Species = gsub(" ","_",Species)) |> 
  mutate(Serration = stringr::str_squish(Serration)) |> 
  as_tibble()

# Phylomorphospace plot.
layout(matrix(c(1:1)),respect = T)
plot(average_pca, axis1 = 1, axis2 = 2,phylo = TRUE, main = NULL, 
     cex = setNames(c(3.5,2.5,1.5),levels(size_species))[size_species[rownames(average_pca$x)]], 
     pch = 21,asp = 2,col = "#2E2D29",
     bg = iucn_colours[iucn_species[rownames(average_pca$x)]],font.lab = 2,cex.axis = .7,cex.lab = .7,
     phylo.par= list(tip.labels = T, node.labels=FALSE, edge.color= "#2E2D29",
                     edge.width= 1, tip.txt.cex =0.5,
                     tip.txt.adj = c(-0.1, -0.1))
     );minor.tick(nx = 2, ny = 2, tick.ratio = 0.5) # IUCN, body size and habitat.

# Legends.
legend("topleft",legend = names(iucn_colours),col = unname(iucn_colours),pch = 19,cex = .5,bty = "n")
# Second legend.
legend("topright",legend = names(size_colours)[c(3,2,1)],col = "black",pch = 21,
       pt.cex = c(1.5,2.5,3.5),
       bty = "n")

# Use "|" for reef asscoiated, "-" for shelf-associated and + for coastal pelagic.
pch_overlay <- 
  ifelse(sorenson_habitats == "Reef", "|",
       ifelse(sorenson_habitats == "Shelf", "-",
              ifelse(sorenson_habitats == "Coastal pelagic", "+", NA)))

habitat_symbol <- factor(setNames(object = pch_overlay, paste(Traits$Genus,Traits$Species, sep = "_")))

par(new = T)
points(average_pca$x[,1:2], pch = as.vector(unname(habitat_symbol[rownames(average_pca$x)])), col = "black", cex = 1.5)
#points(average_pca$x[,c(1,3)], pch = as.vector(unname(habitat_symbol[rownames(average_pca$x)])), col = "black", cex = 1.5)

# Boxplots.
layout(matrix(c(1:2),nrow = 1,ncol = 2),widths = c(4,4),heights = c(7,7),respect = T)
boxplot(average_pca$x[,1] ~ iucn_species[rownames(average_pca$x)],col = "white",border = "black",
        frame.plot=F,boxwex=.8,staplewex=.5,boxlwd=2,medlwd=2,
        ylab = expression(bold("PC1 (78.77%)")),
        xlab = expression(bold("IUCN status")),outcol = scales::alpha("black",.5), outpch = 19,
        notch = FALSE,xaxt="n");abline(h = mean(average_pca$x[,1]),lwd = 2, col = "black", lty = 2)

boxplot(average_pca$x[,2] ~ iucn_species[rownames(average_pca$x)],col = "white",border = "black",
        frame.plot=F,boxwex=.8,staplewex=.5,boxlwd=2,medlwd=2,
        ylab = expression(bold("PC2 (9.31%)")),
        xlab = expression(bold("IUCN status")),outcol = scales::alpha("black",.5), outpch = 19,
        notch = FALSE,xaxt="n");abline(h = mean(average_pca$x[,2]),lwd = 2, col = "black", lty = 2)

# Add text labels for groups on x-axis.
axis(1, at = 1:5, labels = FALSE, tck = -0.02)
text(x = 1:5, y = par()$usr[3] - 0.025*(par()$usr[4] - par()$usr[3]),
     labels = names(iucn_colours), srt = 45, adj = 1, xpd = TRUE)
```

### ANOVA

```{r message=FALSE,results='hide'}
# Prepare centroid size.
average_centroid <- centroid_size
names(average_centroid) <- gsub("_.*","",names(average_centroid))

avg_cz <- 
  data.frame(species = names(average_centroid),
             centroid = unname(average_centroid)
  ) 

ave_cz_transformed <-
  aggregate(avg_cz$centroid,list(avg_cz$species),mean) |> 
  mutate(x = log10(x))

# Create new threat-status factor.
new_threat <- iucn_species[rownames(average_pca$x)] |> 
  as.data.frame() |> 
  setNames(nm = "iucn") |> 
  rownames_to_column(var = "Species") |> 
  mutate(threat = case_when(iucn %in% c("CR","EN","VU") ~ "Threatened", T ~ "Not"))

# Fit models.
iucn_dataframe <- 
  geomorph.data.frame(coords = species_subset,
                      size = unname(setNames(ave_cz_transformed$x,ave_cz_transformed$Group.1)),
                      iucn = as.factor(unname(iucn_species[rownames(average_pca$x)])),
                      # Axes,
                      pc1 = average_pca$x[,1],
                      pc2 = average_pca$x[,2],
                      # Simplified threat status.
                      status = as.factor(new_threat$threat),
                      habitat = as.factor(unname(sorenson_habitats[rownames(average_pca$x)])),
                      body = as.factor(unname(size_species[rownames(average_pca$x)])),
                      tree = carcharhinus_tree
                      )
# Null model.
fit_null <- procD.lm(coords ~ 1,RRPP = T,data = iucn_dataframe, iter = 999)    # Baseline null model.
fit_full <- procD.lm(coords ~ iucn,RRPP = T,data = iucn_dataframe, iter = 999) # Ecological divergence, not accounting for allometry.

fit_axes <- procD.lm(pc1 ~ iucn,RRPP = T,data = iucn_dataframe, iter = 999)
anova(fit_axes)

anova(fit_null,fit_full)
anova(fit_full)

# Pairwise Comparisons.
PW <- pairwise(fit_full, groups = interaction(iucn_dataframe$iucn))

# Threatened versus not-treatened.
fit_iucn <- procD.lm(coords ~ status,RRPP = T,data = iucn_dataframe, iter = 999)
anova(fit_iucn)
```

### Erosion of morphospace

```{r}
# CR species.
cr <- c("Carcharhinus_borneensis","Carcharhinus_hemiodon","Carcharhinus_longimanus","Carcharhinus_porosus")

# Endangered species.
en <- c("Carcharhinus_acronotus","Carcharhinus_amblyrhynchos","Carcharhinus_dussumieri",
        "Carcharhinus_leiodon","Carcharhinus_obscurus","Carcharhinus_perezii",
        "Carcharhinus_plumbeus","Carcharhinus_signatus")

# Vulnerable species.
vu <- c("Carcharhinus_albimarginatus","Carcharhinus_amblyrhynchoides","Carcharhinus_amboinensis",
        "Carcharhinus_brachyurus","Carcharhinus_brevipinna","Carcharhinus_falciformis",
        "Carcharhinus_leucas","Carcharhinus_limbatus",
        "Carcharhinus_melanopterus","Carcharhinus_sealei")

# Near threatened species.
nt <- c("Carcharhinus_altimus","Carcharhinus_isodon","Carcharhinus_macloti","Carcharhinus_sorrah")

# Additive extinction effects.
crEn <- c(cr,en)
crEnVu <- c(cr,en,vu)
crEnVunT <- c(cr,en,vu,nt) # This leaves only LC species alive.

hull.0 <- grDevices::chull(average_pca$x[,1:2]) # Full dental morphospace.
hull.1 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% cr,1:2])
hull.2 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% en,1:2])
hull.3 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% vu,1:2])
hull.4 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% nt,1:2])
hull.5 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% crEn,1:2])
hull.6 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% crEnVu,1:2])
hull.7 <- grDevices::chull(average_pca$x[!rownames(average_pca$x) %in% crEnVunT,1:2])

# Plot extinctions.
# scatterplot3d(average_pca$x[,1:3],
#               cex.symbols = setNames(c(3.5,2.5,1.5),levels(size_species))[size_species[rownames(average_pca$x)]],
#               bg = iucn_colours[iucn_species[rownames(average_pca$x)]],pch = 21
# )

# PC1 vs PC2 vs Depth
depth_akima <- interp(average_pca$x[,1], average_pca$x[,2], 
                      setNames(Traits$Depth,paste(Traits$Genus,Traits$Species, sep = "_"))[rownames(average_pca$x)],
                      linear = T, extrap = F) # Smooth raster grid mappigng depths onto the shape space.

# Plot.
# layout(matrix(c(1:4),ncol = 2,byrow = T),respect = T)

plot(average_pca$x[,1:2], type = "n",
     xlim = c(-0.2022167,0.1361500),
     ylim = c(-0.05479350,0.04883528),
     xlab = "PC 1: 78.97%", ylab = "PC 2: 9.08%", 
     font.lab = 2,cex.axis = .7,cex.lab = .7,);minor.tick(nx = 2, ny = 2, tick.ratio = 0.5)

# Depth profile.
image(depth_akima, add=TRUE,col = colorRampPalette(c("#7f7f7f", "#417865"))(12))
contour(depth_akima, add=TRUE,method = "edge",lwd = .5, col = "white")

# contour(depth_akima, add=TRUE,method = "edge",lwd = .5, col = colorRampPalette(c("#7f7f7f", "#417865"))(12))

# Ranges (minimum convex hulls)
polygon(average_pca$x[,1:2][hull.0,],
        lty = 1,border = "black",lwd = 2) # Full morphospace.

polygon(average_pca$x[!rownames(average_pca$x) %in% cr,1:2][hull.1,],
        lty = 1,border = "black",lwd = 2) # CR eliminated.

# EN eliminated
# polygon(average_pca$x[!rownames(average_pca$x) %in% en,1:2][hull.2,],
#         lty = 1,border = "black",lwd = 2)

# VU eliminated
# polygon(average_pca$x[!rownames(average_pca$x) %in% vu,1:2][hull.3,],
#         lty = 1,border = "black",lwd = 2)

# NT eliminated.
# polygon(average_pca$x[!rownames(average_pca$x) %in% nt,1:2][hull.4,],
#         lty = 1,border = "black",lwd = 2)

# Additive extinctions.
polygon(average_pca$x[!rownames(average_pca$x) %in% crEn,1:2][hull.5,],
        lty = 1,border = "black",lwd = 2) # CR and EN eliminated

polygon(average_pca$x[!rownames(average_pca$x) %in% crEnVu,1:2][hull.6,],
        lty = 1,border = "black",lwd = 2) # CR,EN, and VU eliminated

polygon(average_pca$x[!rownames(average_pca$x) %in% crEnVunT,1:2][hull.7,],
        lty = 1,border = "black",lwd = 2) # CR,EN, VU, and NT eliminated

# Centroids.
c1 <- colMeans(average_pca$x[,c("Comp1", "Comp2")])
c2 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% cr,c("Comp1", "Comp2")])
c3 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% en,c("Comp1", "Comp2")])
c4 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% vu,c("Comp1", "Comp2")])
c5 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% nt,c("Comp1", "Comp2")])

c6 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% crEn,c("Comp1", "Comp2")])
c7 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% crEnVu,c("Comp1", "Comp2")])
c8 <- colMeans(average_pca$x[!rownames(average_pca$x) %in% crEnVunT,c("Comp1", "Comp2")])

# Add centroids
points(c1[1], c1[2], pch = 21, cex = 2, lwd = 2, col = "black", bg = "white")
points(c2[1], c2[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1") # CR removed.
# points(c3[1], c3[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1")
# points(c4[1], c4[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1")
# points(c5[1], c5[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1")
points(c6[1], c6[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1") # CR + EN
points(c7[1], c7[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1") # CR + EN + VU
points(c8[1], c8[2], pch = 21, cex = 2, lwd = 2, col = "black",bg = "#d8cbb1") # CR + EN + VU + NT

# Connecting lines
lines(c(c1[1],c3[1]),c(c1[2],c3[2]), col="black", lwd=2)
lines(c(c1[1],c4[1]),c(c1[2],c4[2]), col="black", lwd=2)
lines(c(c1[1],c5[1]),c(c1[2],c5[2]), col="black", lwd=2)

# Add legend.
# par(mfg = c(3, 2))
# image.plot(depth_akima, col = colorRampPalette(c("#7f7f7f", "#417865"))(12),legend.only = T)
```

### Area loss

```{r results='hide'}
# How much of the volume is impacted? (FA = compute total area and volume) ... Only areas for 2D axes.
# convhulln(average_pca$x[,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% cr,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% en,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% vu,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% nt,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% crEn,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% crEnVu,1:2], options = "FA")$area |> round(digits = 3)
# convhulln(average_pca$x[!rownames(average_pca$x) %in% crEnVunT,1:2], options = "FA")$area |> round(digits = 3)

# Proportion of area lost?
round((0.731 - 0.6527349)/(0.731) * 100,digits = 2) # 10.71%
round((0.731 - 0.652)/(0.731) * 100,digits = 2)     # 10.81%
round((0.731 - 0.572)/(0.731) * 100,digits = 2)     # 21.75%
round((0.731 - 0.282)/(0.731) * 100,digits = 2)     # 61.42%

# Eucldian distance.
euclidean_distance <- 
  function(pt1, pt2) {
    return(sqrt((pt1[1] - pt2[1])^2 + (pt1[2] - pt2[2])^2))
  }

euclidean_distance(pt1 = unname(c1), pt2 = unname(c2))
euclidean_distance(pt1 = unname(c1), pt2 = unname(c3))
euclidean_distance(pt1 = unname(c1), pt2 = unname(c4))
euclidean_distance(pt1 = unname(c1), pt2 = unname(c5))

# Are the shifts statistically significant?
ks.test(average_pca$x[,1],average_pca$x[!rownames(average_pca$x) %in% crEnVu,1])
ks.test(average_pca$x[,2],average_pca$x[!rownames(average_pca$x) %in% crEnVu,2])
```

### Random extinction simulation

```{r}
# Extinction levels.
extinction_levels <- c(4,12,22,26)  # CR, (CR + EN), (CR + EN + VU) and (CR + EN + VU + NT)

q <- 100  

simulated_spaces <- list()

set.seed(1991)

for(k in extinction_levels) {
  for(i in 1:q) {
    remove_species <- sample(rownames(average_pca$x[,1:2]), k)
    reduced_space <- average_pca$x[,1:2][!rownames(average_pca$x[,1:2]) %in% remove_species, ]
    
    simulated_spaces[[paste0("remove_", k, "_sim_", i)]] <- reduced_space
  }
}

extinction_results <-
  bind_rows(lapply(simulated_spaces, colMeans), .id = "scenario") |> 
  mutate(extinction = sub("_sim_.*", "", scenario)) |> 
  mutate(extinction = factor(extinction, levels = c("remove_4","remove_12","remove_22","remove_26"))) |> 
  as.data.frame()

# Compute summary statistics based on PC1 and PC2.
first_axis <- 
  extinction_results |> 
  group_by(extinction) |> 
  summarise(mean = intervals(Comp1)$mean,
            lwr_ci = intervals(Comp1)$Lower.CI,
            upr_ci = intervals(Comp1)$Upper.CI,
            var = "PC1")

second_axis <- 
  extinction_results |> 
  group_by(extinction) |> 
  summarise(mean = intervals(Comp2)$mean,
            lwr_ci = intervals(Comp2)$Lower.CI,
            upr_ci = intervals(Comp2)$Upper.CI,
            var = "PC2")

# Plot results.
combined_results <- rbind(first_axis,second_axis)

# layout(matrix(c(1,2,3,3),nrow = 2,ncol = 2),
#        widths = c(6,6,7),heights = c(7,7,7),
#        respect = T)

layout(matrix(c(1:2),nrow = 1,ncol = 2),respect = T)

# par(mar = c(4,5 2,1))
# layout(matrix(c(1,3,2,3), 2, 2, byrow = TRUE),respect = T)
plot(combined_results$mean[1:4], 
     ylim = range(c(combined_results$lwr_ci[1:4], 0.08713091)), 
     xlab = "Extinction Scenario",
     ylab = expression(bar(PC1)), 
     xaxt = "n",
     bg = "white",font.lab = 2,
     pch = 19)

# The observed changes in mean value for PC1 and PC2 under each extinction scenario
points(1:4,c(c2[1],c6[1],c7[1],c8[1]),col = "#2659A4", pch = 19)
# Add solid line for reference.
abline(h = mean(average_pca$x[,1]),col = "#d8cbb1",lwd = 2)

# Add error bars
arrows(x0 = 1:4, y0 = combined_results$lwr_ci[1:4],
       x1 = 1:4, y1 = combined_results$upr_ci[1:4],
       angle = 90, code = 3, length = 0.05,lty = 1)

# PC2.
plot(combined_results$mean[5:8], 
     ylim = range(c(combined_results$lwr_ci[5:8], 0.0081644476)), 
     xlab = "Extinction Scenario", 
     ylab = expression(bar(PC2)),font.lab = 2,
     bg = "white",
     pch = 19)

# The observed changes in mean value for PC2 under each extinction scenario
points(1:4,c(c2[2],c6[2],c7[2],c8[2]),col = "#2659A4", pch = 19)

abline(h = mean(average_pca$x[,2]),col = "#d8cbb1",lwd = 2)

arrows(x0 = 1:4, y0 = combined_results$lwr_ci[5:8],
       x1 = 1:4, y1 = combined_results$upr_ci[5:8],
       angle = 90, code = 3, length = 0.05)
```

### Disparity analysis

```{r}
# Plot species richness.
# par(mar = c(5,5,3,4))
plot(x = 1:5, y = c(30,26,18,8,4),ylab = "Species richness",
     pch = 19,type = "b",xlab = "Extinction scenario",
     font.lab = 2);abline(h = 30,lwd = 2,lty = 2)

# Plot pairwise Procrustes distance.
par(new = TRUE)

plot(c(
  mean(as.matrix(dist(two.d.array(A = species_subset)))),
  mean(as.matrix(dist(two.d.array(A = species_subset[,,!dimnames(species_subset)[[3]] %in% cr])))),
  mean(as.matrix(dist(two.d.array(A = species_subset[,,!dimnames(species_subset)[[3]] %in% crEn])))),
  mean(as.matrix(dist(two.d.array(A = species_subset[,,!dimnames(species_subset)[[3]] %in% crEnVu])))),
  mean(as.matrix(dist(two.d.array(A = species_subset[,,!dimnames(species_subset)[[3]] %in% crEnVunT]))))
  ), 
  ylim = c(0.07,0.12),
  xlab = "", ylab = "",type = "p",pch = 17, cex = 1,
  col = c("black","#B71234","#B71234","#B71234","#B71234"),
  xaxt = "n", yaxt = "n")

axis(side = 4)
mtext("Mean Pairwise Procrustes distance", side = 4, line = 3)

# Simulate extinction at random.
simulated_disparity <- list()

set.seed(1992)

for(k in extinction_levels) {
  level_results <- vector("list", q)
  for(i in 1:q) {
    eS <- sample(dimnames(species_subset)[[3]], k)
    eP <- species_subset[,,!dimnames(species_subset)[[3]] %in% eS]
    level_results[[i]] <- eP
  }
  simulated_disparity[[as.character(k)]] <- level_results
}

# Four species extinct at random
mean_results_1 <- vector(mode = "list")

for(i in seq_along(simulated_disparity[[1]])) {
  tr1 <- mean(as.matrix(dist(two.d.array(A = simulated_disparity[[1]][[i]]))))
  mean_results_1[[i]] <- tr1
  mean_results_1 <- unlist(mean_results_1)
}

# Twelve species extinct at random
mean_results_2 <- vector(mode = "list")

for(i in seq_along(simulated_disparity[[2]])) {
  tr2 <- mean(as.matrix(dist(two.d.array(A = simulated_disparity[[2]][[i]]))))
  mean_results_2[[i]] <- tr2
  mean_results_2 <- unlist(mean_results_2)
}

# 22 species extinct at random
mean_results_3 <- vector(mode = "list")

for(i in seq_along(simulated_disparity[[3]])) {
  tr3 <- mean(as.matrix(dist(two.d.array(A = simulated_disparity[[3]][[i]]))))
  mean_results_3[[i]] <- tr3
  mean_results_3 <- unlist(mean_results_3)
}

mean_results_4 <- vector(mode = "list")

for(i in seq_along(simulated_disparity[[4]])) {
  tr4 <- mean(as.matrix(dist(two.d.array(A = simulated_disparity[[4]][[i]]))))
  mean_results_4[[i]] <- tr4
  mean_results_4 <- unlist(mean_results_4)
}

# Combine results.
combined_ApPD_res <- 
  cbind(mean_results_1,mean_results_2,mean_results_3,mean_results_4) |> 
  as.data.frame() |> 
  setNames(nm = c("remove_4","remove_12","remove_22","remove_26")) |> 
  pivot_longer(cols = c(1:4),names_to = "extinction")

APPD_res <- 
  combined_ApPD_res |> 
  group_by(extinction) |> 
  summarise(mean = intervals(value)$mean,
            lwr_ci = intervals(value)$Lower.CI,
            upr_ci = intervals(value)$Upper.CI
            ) |> 
  ungroup() |> 
  mutate(extinction = factor(extinction, levels = c("remove_4","remove_12","remove_22","remove_26"))) |> 
  add_row(extinction = "remove_none", mean = NA, lwr_ci = NA, upr_ci = NA)

# par(new = TRUE)
# plot(APPD_res$mean[c(5,4,1:3)],
#      ylim = range(c(APPD_res$lwr_ci, APPD_res$upr_ci),na.rm = T), 
#      xlab = "", ylab = "",type = "p",pch = 17, cex = 1,
#      col = "lightblue",
#      xaxt = "n", yaxt = "n")

arrows(
  x0 = 1:5, y0 = APPD_res$lwr_ci[c(5,4,1:3)],
  x1 = 1:5, y1 = APPD_res$upr_ci[c(5,4,1:3)],
  angle = 90, code = 3, length = 0.05,col = "#009AB4",
)

points(APPD_res$mean[c(5,4,1:3)],pch = 17, col = "#009AB4")

# Add legend.
legend("bottomleft",legend = c("Species Diveristy","APPD","Randomized-APPD"),
       col = c("black","#B71234","#009AB4"),
       pch = c(19,17,17),cex = .8)

#axis(side = 4, line = 4)
#mtext("Simulated Extinction", side = 4, line = 6)
```

## Sensitivity analysis

```{r}
# Species average based on upper and lower teeth.
jaw_coords <- tooth_gpa$coords

dimnames(jaw_coords)[[3]] <- gsub("-"," ",dimnames(jaw_coords)[[3]])
dimnames(jaw_coords)[[3]] <- gsub("_U", "_Upper",dimnames(jaw_coords)[[3]])
dimnames(jaw_coords)[[3]] <- gsub("_L", "_Lower",dimnames(jaw_coords)[[3]])
dimnames(jaw_coords)[[3]] <- gsub("[0-9]+", "", dimnames(jaw_coords)[[3]])

# Computing an average shape.
jaw_subset <- coords.subset(A = jaw_coords, group = as.factor(dimnames(jaw_coords)[[3]]))

jaw_subset <- 
  lapply(jaw_subset, mshape) |>
  simplify2array()

dim(jaw_subset) # 60 configurations.

# Final pre-processing.
upper_jaw_coords <- jaw_subset[,,grep(pattern = "Upper",x = dimnames(jaw_subset)[[3]],value = T)]
lower_jaw_coords <- jaw_subset[,,grep(pattern = "Lower",x = dimnames(jaw_subset)[[3]],value = T)]

dimnames(upper_jaw_coords)[[3]] <- gsub("_.*","",x = dimnames(upper_jaw_coords)[[3]])
dimnames(upper_jaw_coords)[[3]] <- gsub(" ","_", x = dimnames(upper_jaw_coords)[[3]])

dimnames(lower_jaw_coords)[[3]] <- gsub("_.*","",x = dimnames(lower_jaw_coords)[[3]])
dimnames(lower_jaw_coords)[[3]] <- gsub(" ","_", x = dimnames(lower_jaw_coords)[[3]])

# PCA.
average_upper_jaw <- gm.prcomp(A = upper_jaw_coords,phy = carcharhinus_tree)
average_lower_jaw <- gm.prcomp(A = lower_jaw_coords,phy = carcharhinus_tree)

# Serration type for upper teeth by species.
upper_serrations_frame <- 
  Data |> 
  filter(Jaw == "Upper") |> 
  select(Species,Serration) |> 
  distinct(Species,.keep_all = T) |> 
  mutate(Species = gsub(" ","_",Species)) |> 
  mutate(Serration = stringr::str_squish(Serration)) |> 
  as_tibble()

lower_serrations_frame <- 
  Data |> 
  filter(Jaw == "Lower") |> 
  select(Species,Serration) |> 
  distinct(Species,.keep_all = T) |> 
  mutate(Species = gsub(" ","_",Species)) |> 
  mutate(Serration = stringr::str_squish(Serration)) |> 
  as_tibble()

# Plot upper dental configurations.
layout(matrix(c(1:1)),respect = T)
plot(average_upper_jaw, axis1 = 1, axis2 = 2,phylo = TRUE, main = NULL, 
     cex = setNames(c(3.5,2.5,1.5),levels(size_species))[size_species[rownames(average_upper_jaw$x)]], 
     pch = 21,asp = 2,col = "#2E2D29",
     bg = iucn_colours[iucn_species[rownames(average_upper_jaw$x)]],font.lab = 2,cex.axis = .7,cex.lab = .7,
     phylo.par= list(tip.labels = T, node.labels=FALSE, edge.color= "#2E2D29",
                     edge.width= 1, tip.txt.cex =0.5,
                     tip.txt.adj = c(-0.1, -0.1))
     ); minor.tick(nx = 2, ny = 2, tick.ratio = 0.5)

par(new = T)
points(average_upper_jaw$x[,1:2], pch = as.vector(unname(habitat_symbol[rownames(average_upper_jaw$x)])), col = "black", cex = 1.5)

plot(average_lower_jaw, axis1 = 1, axis2 = 2,phylo = TRUE, main = NULL,
     # Serrations.
     # lwd =c(4,2,1,.5)[as.factor(setNames(lower_serrations_frame$Serration,lower_serrations_frame$Species))],
     # Body size.
     cex = setNames(c(3.5,2.5,1.5),levels(size_species))[size_species[rownames(average_pca$x)]], 
     pch = 21,asp = 2,col = "#2E2D29",
     # IUCN status.
     bg = iucn_colours[iucn_species[rownames(average_pca$x)]],font.lab = 2,cex.axis = .7,cex.lab = .7,
     phylo.par= list(tip.labels = F, node.labels=FALSE, edge.color= "#2E2D29",
                     edge.width= 1, tip.txt.cex =0.5,
                     tip.txt.adj = c(-0.1, -0.1))
     ) 

par(new = T)
points(average_lower_jaw$x[,1:2], pch = as.vector(unname(habitat_symbol[rownames(average_lower_jaw$x)])), col = "black", cex = 1.5)
dev.off()

# par(mfrow = c(1,2))
# plotRefToTarget(average_upper_jaw$shapes$shapes.comp1$min,average_upper_jaw$shapes$shapes.comp1$min)
# plotRefToTarget(average_upper_jaw$shapes$shapes.comp1$max,average_upper_jaw$shapes$shapes.comp1$max)
# plotRefToTarget(average_upper_jaw$shapes$shapes.comp2$min,average_upper_jaw$shapes$shapes.comp2$min)
# plotRefToTarget(average_upper_jaw$shapes$shapes.comp2$max,average_upper_jaw$shapes$shapes.comp2$max)
# 
# plotRefToTarget(average_lower_jaw$shapes$shapes.comp1$min,average_lower_jaw$shapes$shapes.comp1$min)
# plotRefToTarget(average_lower_jaw$shapes$shapes.comp1$max,average_lower_jaw$shapes$shapes.comp1$max)
# plotRefToTarget(average_lower_jaw$shapes$shapes.comp2$min,average_lower_jaw$shapes$shapes.comp2$min)
# plotRefToTarget(average_lower_jaw$shapes$shapes.comp2$max,average_lower_jaw$shapes$shapes.comp2$max)
# dev.off()
```

## Phylogenetic signal

```{r}
# Test for phylogenetic signal in tooth shape.
carcharhinus_signal <- geomorph::physignal(A = species_subset,phy = carcharhinus_tree,iter = 999,seed = 1)
# carcharhinus_signal

# Effect size.
carcharhinus_effect <- geomorph::physignal.z(A = species_subset,phy = carcharhinus_tree)
# carcharhinus_effect
```

## Ecological diversity

```{r}
d1 <- vegdist(x = as.matrix(unclass(clr(acomp(closed_diets)))[,1:10]), method = "euclidean") # All.
# Re-caluclate after elimination of CR, EN, VU, and NT.
d2 <- vegdist(x = as.matrix(unclass(clr(acomp(closed_diets[!rownames(closed_diets) %in% cr,])))[,1:10]), method = "euclidean") # CR
d3 <- vegdist(x = as.matrix(unclass(clr(acomp(closed_diets[!rownames(closed_diets) %in% crEn,])))[,1:10]), method = "euclidean") # CR + EN
d4 <- vegdist(x = as.matrix(unclass(clr(acomp(closed_diets[!rownames(closed_diets) %in% crEnVu,])))[,1:10]), method = "euclidean") # CR + EN + VU
d5 <- vegdist(x = as.matrix(unclass(clr(acomp(closed_diets[!rownames(closed_diets) %in% crEnVunT,])))[,1:10]), method = "euclidean") # CR + EN + VU + NT

# Multivariate homogeneity of groups dispersions.
bd0 <- betadisper(d1, group = Diets$IUCN_2024)

ex0 <- betadisper(d1, group = as.factor(rep("A",27)))
ex1 <- betadisper(d2, group = as.factor(rep("A",25)))
ex2 <- betadisper(d3, group = as.factor(rep("A",18)))
ex3 <- betadisper(d4, group = as.factor(rep("A",8)))
ex4 <- betadisper(d5, group = as.factor(rep("A",4)))

bd0$distances |> 
  as.data.frame() |> 
  setNames(nm = "distance") |> 
  rownames_to_column(var = "Species") |> 
  left_join(Traits |> select(Genus, Species, IUCN,Size_Subclass) |> mutate(Species = paste(Genus,Species,sep = "_")),by = "Species") |> 
  mutate(IUCN = factor(IUCN, levels = c("CR","EN","VU","NT","LC"))) |> 
  ggplot(mapping = aes(x = IUCN, y = distance)) +
  geom_boxplot(outlier.shape = NA, width=.8) +
  geom_jitter(aes(colour = Size_Subclass)) +
  theme_bw() +
  theme(legend.position = "top",aspect.ratio = 1) # measuring directional change in ecological space.

# Plot diet results.
layout(matrix(c(1,1,1,2,3,4), nrow = 2, byrow = TRUE),
       widths = c(3,3,3), 
       heights = c(6,4),
       respect = TRUE)

par(mar = c(2,4,4,2))
phylo.heatmap(tree = tree_diets, X = Diets[,6:15], standardize = F,
              colors = rev(sequential_hcl(11,palette = "Red-Yellow")),split=c(0.6,0.4),
              fsize = .5) # Overwheling fish-eaters across the board.

# ODB.
par(mar = c(5, 4, 2, 1))
plot(x = odb_results$Index,y = odb_results$Axis,
     cex = setNames(c(3.5,2.5,1.5),levels(size_species))[size_species[odb_results$Species]], 
     pch = 21,col = "#2E2D29",
     bg = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50")[iucn_categories],
     font.lab = 2,cex.axis = .7,cex.lab = .7)

par(new = T)
points(x = odb_results$Index,y = odb_results$Axis, pch = as.vector(unname(habitat_symbol[odb_results$Species])),
       col = "black", cex = 1.5)
# Add quantiles.
abline(h = c(quantile(odb_results$Axis, c(0.33, 0.66))),col = "grey50", lty = 2)

# text(x = odb_results$Index,y = odb_results$Axis,labels = odb_results$Species,cex = .5)

boxplot(bd0, col = "white",border = c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50"),
        boxwex=.8,staplewex=.5,boxlwd=2,medlwd=2)

points(jitter(as.numeric(bd0$group)), bd0$distances, 
       col = scales::alpha(c("#B71234","#E98300","#FDC82F","#A0A831","#4CAF50"),.5)[bd0$group],
       pch = 19)

# bd0$distances |> 
#   as.data.frame() |> 
#   setNames(nm = "value") |> 
#   rownames_to_column(var = "Species") |> 
#   left_join(Traits |> select(Genus,Species, IUCN) |> mutate(Species = paste(Genus,Species,sep = "_")), by = "Species") |> 
#   filter(IUCN == "VU") |> 
#   arrange(desc(value))

mean_dispersion <-
  c(mean(ex0$distances),
    mean(ex1$distances),
    mean(ex2$distances),
    mean(ex3$distances),
    mean(ex4$distances)
    )

plot(x = 1:5,y = mean_dispersion,type = "o",lwd = 2,ylim = c(0,7),
     pch = 19, ylab = "Mean ecological dispersion",
     font.lab = 2)

grp_sd <- c(
  sd(ex0$distances),
  sd(ex1$distances),
  sd(ex2$distances),
  sd(ex3$distances),
  sd(ex4$distances)
)

arrows(
  x0 = 1:5, y0 = mean_dispersion - grp_sd,
  x1 = 1:5, y1 = mean_dispersion + grp_sd,
  angle = 90, code = 3, length = 0.05, lwd = 1.5
)

# Is there a relationship between IUCN status and ODB values?
# summary(aov(odb_results$Axis ~ odb_results$IUCN)) # No.
# summary(aov(odb_results$Axis ~ odb_results$Sorensen_habitat)) # Habitat, no.
# summary(aov(odb_results$Axis ~ odb_results$Size_Subclass)) # Body size, yes. Which one? Large versus small sharks have difference values.

# TukeyHSD(aov(odb_results$Axis ~ odb_results$Size_Subclass), conf.level=.95)

# PERMANOVA.
# adonis2(dist(bd0$distances) ~ bd0$group)
```

## Download data

{{< downloadthis Carcharhinus.xlsx dname="Carcharhinus information dataset" label="Download Carcharhinus data used in this study as .xlsx" icon="file-earmark-spreadsheet" type="success" >}}
{{< downloadthis Diets.xlsx dname="Carcharhinus diets" label="Download Carcharhinus diets as .xlsx" icon="file-earmark-spreadsheet" type="success" >}}
{{< downloadthis Traits.xlsx dname="Carcharhinus traits" label="Download Carcharhinus traits as .xlsx" icon="file-earmark-spreadsheet" type="success" >}}
